Untitled

### 2024-10-07 analysis/paper/access-ice.qmd
Sentence wrap

---
title: "Untitled"
format: gfm
execute: 
  echo: false
  warning: false
  message: false
  cache: false
filters:
  - ../../_extensions/pandoc-ext/abstract-section/abstract-section.lua
  - ../../_extensions/mloubout/critic-markup/critic-markup.lua
  - ../../_extensions/ute/search-replace/search-replace.lua
search-replace:
  +S2: ACCESS-S2
  +S1: ACCESS-S1
  +CDR: CDR
bibliography: references.bib
---

```{r setup, include=FALSE}
library(lubridate)
library(ggplot2)
library(data.table)
library(metR)
library(rcdo)
library(tagger)
library(patchwork)

@@

+S2 is the next saeasonal prediction system after ACCESS-S1. It uses virtually the same exact model configuration except for some model versions being different. The main difference are the initial conditions. While S1 used initial conditions from the UK Met Office, S2 uses the Bureau's own system. {== Description of the system ==}{>>Atmosphere and ocean. Highlight no ice data assimilation. Highligh that the hope is that ice will respond to SST/atmosphere DA<<}

@@

There is not a lot of data on sea ice properties, especially for things like thickness, age, etc. However there are relativelly reliable satellite-derived estimates of sea ice concentration, which esetimates the proportion of each grid area that is covered with ice. These products are not perfect and there still exists a fair bit of observation al uncertainty. To account for this uncertanty we use multiple datasets.

@@

Comiso, J.C., Gersten, R.A. 2023. Bootstrap Sea Ice Concentrations from Nimbus-7 SMMR and DMSP SSM/I-SSMIS, Version 4. \[Indicate subset used\]. Boulder, Colorado USA. NASA National Snow and Ice Data Center Distributed Active Archive Center. https://doi.org/10.5067/X5LG68MH013O. \[Date Accessed\] {\>\>Add to zotero\<\<}

@@

NOAA/NSIDC's Climate Data Record V4 combines the Bootstrap and NASA Team estimates in an attempt to overcome each algorithm's weaknesses. {\>\>More detail on how are they combined and their properties.\<\<} As such, it is is not a trully independend dataset.

@@

{\>\>Why is ERA5 better or worse than satellites?\<\<}

@@

FOr evaluation purposes, we use a series of measures.

@@

We also compute the Integrated Ice Edge Error (IIEE) {\>\>CITE\<\<}. This is defined as the area in which the model misspredicts the presence of more than 15% ice. That is, dichotomise sea ice concentration into areas with more and less than 15% sea ice both in the forecast and observations; the IIEE is the area in which forecast and observations differ.

@@

@fig-hindcast-extent a shows the seasonal cycle of Sea Ice Extent for the +S2 hindcast and `r length(datasets)-1` observational datasets. 
+S2 shows a severe low extent bias, especially in the late summer-early autumn. 
This is due primarily to a faster and longer melt season between January and March and slower growth during March and April 
This is then balanced with faster growth between May and July ([Fig. @fig-mean-growth]).
This bias is common in climate models [ref!] and it's been linked to xxxx??

Comparing +S2 with +S1, the latter has a smaller bias, especially at low lags ([Fig. @fig-hindcast-extent b]) even though the typical growth rates are very similar between both models ([Fig. @fig-mean-growth]).
At larger lags, +S1's bias in summer and autumn is very similar to +S2`s.
This suggests that this lower sea ice state is closer to the models' equilibrium, indicating that it is an issue with model formulation that was being corrected by the data assimilation system in +S1.

@@

climatologies |> 
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(aes(z = S2 - cdr, fill = after_stat(level)), 
                    breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)) +
  scale_fill_divergent_discretised("ACCESS-S2 bias") +
  geomcoord_antarctica +
  geom_antarctica_fill +
  facet_wrap(~ lubridate::month(time, label = TRUE)) +
  wide_legend +
  tag_facets()
```
@fig-bias shows the difference in monthly mean sea ice concentrations between +CDR and +S2 reanalysis. 
From October to May, the model underestimates sea ice concentrations pretty much everywhere there is ice except for the deep Weddell Sea in April and May, where sea ice concentrations saturate to 1. 
In winter, the differences are mostly on the sea ice edge, with slight positive bias in XXX and negative bias around the Indian Ocean sector.

@@

{\>\>Add reginoal extents?\<\<}

@@

@fig-rmse shows the median and 95% range of RMSE of sea ice concentration anomalies for +S2 forecasts compared with a benchmark of persistence. 
This figure uses +CDR data, but the results are nearly identical compared with ERA5 or Bootstrap. 
Due to errors in the initial conditions, it is expected that a persistence forecast would be better than the model forecast at very short lags, but that the persistence forecast errors would grow faster and eventually surpass the mode forecast, at which time is statistically useful {==I've got this from CC at the ICTP summer school and makes sense, but it would be great to have a referece?==}.
Here the persistence errors are almost always lower than the +S2 forecast, indicating that the model doesn't have skill at any lag and in any month. 
The only exception is the RMSE around Febraury forecasted from June onwards.

{==Add regional?==}

>>>>

---
title: "Untitled"
format: gfm
execute: 
  echo: false
  warning: false
  message: false
  cache: false
filters:
  - ../../_extensions/pandoc-ext/abstract-section/abstract-section.lua
  - ../../_extensions/mloubout/critic-markup/critic-markup.lua
  - ../../_extensions/ute/search-replace/search-replace.lua
search-replace:
  +S2: ACCESS-S2
  +S1: ACCESS-S1
  +CDR: CDR
bibliography: references.bib
editor: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
library(lubridate)
library(ggplot2)
library(data.table)
library(metR)
library(rcdo)
library(tagger)

@@

+S2 is the next saeasonal prediction system after ACCESS-S1.
It uses virtually the same exact model configuration except for some model versions being different.
The main difference are the initial conditions.
While S1 used initial conditions from the UK Met Office, S2 uses the Bureau's own system.
{== Description of the system ==}{\>\>Atmosphere and ocean.
Highlight no ice data assimilation.
Highligh that the hope is that ice will respond to SST/atmosphere DA\<\<}

@@

There is not a lot of data on sea ice properties, especially for things like thickness, age, etc.
However there are relatively reliable satellite-derived estimates of sea ice concentration, which estimates the proportion of each grid area that is covered with ice.
These products are not perfect and there still exists a fair bit of observation al uncertainty.
To account for this uncertainty we use multiple datasets.

@@

@@

NOAA/NSIDC's Climate Data Record V4 combines the Bootstrap and NASA Team estimates in an attempt to overcome each algorithm's weaknesses.
{\>\>More detail on how are they combined and their properties.\<\<} As such, it is is not a truly independent dataset.

@@

{\>\>Why is ERA5 better or worse than satellites?\<
\<}

@@

For evaluation purposes, we use a series of measures.

@@

We also compute the Integrated Ice Edge Error (IIEE) [@goessling2016].
This is defined as the area in which the model misspredicts sea ice concentration being above or belog 15% ice.
That is, dichotomise sea ice concentration into areas with more and less than 15% sea ice both in the forecast and observations; the IIEE is the area in which forecast and observations differ.

@@

@fig-hindcast-extent a shows the seasonal cycle of Sea Ice Extent for the +S2 hindcast and `r length(datasets)-1` observational datasets.
+S2 shows a severe low extent bias, especially in the late summer-early autumn.
This is due primarily to a faster and longer melt season between January and March and slower growth during March and April This is then balanced with faster growth between May and July ([Fig. @fig-mean-growth]).
This bias is common in climate models \[ref!\]
and it's been linked to xxxx??

Comparing +S2 with +S1, the latter has a smaller bias, especially at low lags ([Fig. @fig-hindcast-extent b]) even though the typical growth rates are very similar between both models ([Fig. @fig-mean-growth]).
At larger lags, +S1's bias in summer and autumn is very similar to +S2\`s.
This suggests that this lower sea ice state is closer to the models' equilibrium, indicating that it is an issue with model formulation that was being corrected by the data assimilation system in +S1.

@@

climatologies |> 
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(aes(z = S2 - cdr, fill = after_stat(level)), 
                    breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)) +
  scale_fill_divergent_discretised("ACCESS-S2 bias") +
  geomcoord_antarctica +
  geom_antarctica_fill +
  facet_wrap(~ lubridate::month(time, label = TRUE)) +
  wide_legend +
  tag_facets()
```

@fig-bias shows the difference in monthly mean sea ice concentrations between +CDR and +S2 reanalysis.
From October to May, the model underestimates sea ice concentrations pretty much everywhere there is ice except for the deep Weddell Sea in April and May, where sea ice concentrations saturate to 1.
In winter, the differences are mostly on the sea ice edge, with slight positive bias in XXX and negative bias around the Indian Ocean sector.

@@

{\>\>Add reginoal extents?\<
\<}

@@

@fig-rmse shows the median and 95% range of RMSE of sea ice concentration anomalies for +S2 forecasts compared with a benchmark of persistence.
This figure uses +CDR data, but the results are nearly identical compared with ERA5 or Bootstrap.
Due to errors in the initial conditions, it is expected that a persistence forecast would be better than the model forecast at very short lags, but that the persistence forecast errors would grow faster and eventually surpass the mode forecast, at which time is statistically useful {==I've got this from CC at the ICTP summer school and makes sense, but it would be great to have a referece?=
=}.
Here the persistence errors are almost always lower than the +S2 forecast, indicating that the model doesn't have skill at any lag and in any month.
The only exception is the RMSE around Febraury forecasted from June onwards.

{==Add regional?=
=}
